---
title: "p8105_hw2_fs2968"
output: github_document
date: "2026-09-30"
---
Importing necessary libraries
```{r}
library(tidyverse)
```


## Problem 1, NYC Transit Data

```{r}
t_data = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude, starts_with("route"), entry, vending, entrance_type, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```


In the above code, we read in the NYC transit subway data-set that includes information about the following points: line, station name, station latitude and longitude, the route it lies on, whether there is an entry and whether there is a fee, and the entrance type. To clean the data, we first use the `clean_names` function from the janitor package to ensure all of the column names are cleaned so that there are no dots, uppercase, and overall more consistent. The next step was selecting all of the columns that we want to keep using the `select` function applied on the imported data. Note, since we wanted all route information, we use the `starts_with` function to include all columns with names that start with route to simplify the code. Lastly, we use the `mutate` function to change the data-set, modifying the entry column to be a logical vector with `TRUE` and `FALSE` instead of characters `"YES"` and `"NO"`. In the import, you could consider declaring `NA= c("NA", "", ".")` a cleaning step, but it is more of an identification for the read_csv function of what the missing values will be. 

The post-clean data set has dim (`r dim(t_data)`), which represents `r dim(t_data)[1]` transit stations (rows) with `r dim(t_data)[2]`  descriptor fields per row (columns). After all of the cleaning, the data is mostly tidy. The names look good and data typing is in line with normal standards, but one thing we are missing is that some of the column names are observation values, specifically route1,...,route11. Instead of separating these into 11 different groups, we should instead have this as one column for routes.

To find the number of distinct stations, which vary by name and line, we can use the `distinct` and `nrow` functions:

```{r}
nrow(distinct(t_data, station_name, line))
```

To find the number of stations that are ADA compliant, we look at all stations that are ADA compliant, and then add the `distinct` function to ensure we aren't double counting any stations, and then use `nrow` to find the total.

```{r}
filter(t_data, ada == TRUE) |> 
  distinct(station_name, line) |> 
  nrow()
```

To find the number of station entrances/exits without vending that allow entrance, we first `filter` t_data for those that allow entrance and those without vending. Once we have that filtered data-set, we similarly use `distinct` to make sure we aren't repeating stations and use `nrow` to find the total. Now we have all of the pieces and we simply divide that value by the total number of distinct stations without vending to find the proportion:

```{r}
numerator = filter(t_data, entry == TRUE, vending == "NO") |> 
            distinct(station_name, line) |> 
            nrow()

denominator = filter(t_data, vending == "NO") |> 
              distinct(station_name, line) |> 
              nrow()

numerator/denominator
```


To reformat the data, we will use the `pivot_longer` function to change the route columns to describe the route number as one column and the route name as another. Before we do that, we first need to change the route names to all be char because route 1,2,3,4,... are default to doubles:

```{r}
reformatted_t_data = 
  t_data |> 
  mutate(
    across(starts_with("route"), as.character)
  ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )

```
Now we can answer the find the number of distinct stations that serve the A train by using the `filter` function to only look at stations that serve the A, and then use the `distinct` and `nrow` functions to find the total distinct stations:

```{r}
filter(reformatted_t_data, route_name == "A") |> 
  distinct(station_name, line) |> 
  nrow()
```

To further find the number that are ADA compliant, we simply add that to the previous `filter` function:

```{r}
filter(reformatted_t_data, route_name == "A", ada == TRUE) |> 
  distinct(station_name, line) |> 
  nrow()
```


## Problem 2, Trash Wheel Dataset

We first read in the Trash Wheel Data-set and remove non-data entries and not including dumpster specific data. We also can adjust the names to be more reasonable and round the number of sports balls to the nearest integer and convert to an integer type using `as.integer`.

```{r}
trash_data = 
  readxl::read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx", 
    sheet="Mr. Trash Wheel", 
    range="A2:N742",
    na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  mutate(
    across(sports_balls, round),
    across(sports_balls, as.integer),
    owner = "Mr. Trash",
    across(year, as.double)
  )

```
Now we can grab the other two datasets and clean them using a similar process:

```{r}
prof_data = 
  readxl::read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx", 
    sheet="Professor Trash Wheel", 
    range="A2:M139",
    na = c("NA", "", ".", "dive"))|> 
  janitor::clean_names() |> 
  mutate(
    owner = "Professor Trash",
    )

gwyn_data = 
  readxl::read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx", 
    sheet="Gwynnda the Good Wheel of the W", 
    range="A2:L394",
    na = c("NA", "", "."))|> 
  janitor::clean_names() |> 
  mutate(owner = "Gwynns Trash")

```


Now we can use the `bind_rows` function to join these data-sets together. Before that, we can create a column in all three called "owner", which designates who the data comes from, editted in the above code blocks. 

```{r}
total_trash = bind_rows(trash_data, prof_data, gwyn_data)
```

The resulting data-set `total_trash` is the combination of all three excel sheets. This combined data-set includes `r nrow(total_trash)` observations (rows) and `r ncol(total_trash)` variables (columns). There are both identification variables and descriptive variables. The identification variables include `r names(total_trash)[0:4]`, which help us identify the dumpsters and when the were observed. The descriptive variables include `r names(total_trash)[5:ncol(total_trash)]`, which describe the weight, volume, and breakdown of what is found in the dumpster. The weight is measured in tons with mean value of `r mean(pull(total_trash, weight_tons))`, and the volume is measured in cubic yards with an mean value of `r mean(pull(total_trash, volume_cubic_yards)) `. The other descriptive variables that detail what actual items are in the dumpters use integer interpretations, for example, the first 3 dumpsters had `r pull(total_trash, plastic_bags)[0:3]` plastic bags.

For the available data, the total weight of trash collected by Professor Trash Wheel was `r sum(pull(total_trash, weight_tons))` tons. We can additionally find the total number of cigarette butts collected by Gwynnda in June of 2022 as `r sum(pull(filter(total_trash, owner == "Gwynns Trash", month == "June", year == 2022), cigarette_butts))`.


## Problem 3, Alzheimer's Disease Dataset 

First, we need to import, clean, and tidy the dataset.
```{r}
a_data = 
  read_csv("data/data_mci/MCI_baseline.csv",
           na = c("NA", ".", ""),
           skip= 1
           ) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex, 
      1 ~ "Male",
      0 ~ "Female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ TRUE,
      0 ~ FALSE
    )
  )
#We don't include the filter in the a_data because we need to describe the number of people filtered out later
a_valid_data = filter(a_data, !current_age > age_at_onset | is.na(age_at_onset))

```

We import our dataset with the usual `read_csv` function and mark "NA", ".", and "" as the potential na values. Taking a look at the data, it seems like . is the used na value. Next, we use clean names to make sure the names are consistent and easy to use, then we use the mutate function to change the sex variable to characters "Male" and "Female" and change the APOE4 gene variables to `TRUE` and `FALSE` since it designates whether that gene is present. Lastly, we removed any participants who had an age of onset prior to the baseline using the `filter` function. Note, we include those with na values for age of onset since those are participants who were in the study to not develop MCI during the study.

For relevant features of the dataset, there are `r nrow(a_valid_data)` valid participants (rows) and `r ncol(a_valid_data)` variables per participant. These variables include `r names(a_valid_data)`, which help describe participants risk of developing MCI given age, education, sex, and presence of the APEO4 gene. In total, there were `r nrow(a_data)` participants recruited, but since `r nrow(filter(a_data, current_age > age_at_onset))` participants had MCI at baseline, we only include `r nrow(a_valid_data)` participants in our analysis. Of these valid participants, `r nrow(filter(a_valid_data, !is.na(age_at_onset)))` developed MCI during the study and follow-up period. The average baseline age is `r mean(pull(a_valid_data, current_age))` and `r (nrow(filter(a_valid_data, sex == "Female")) / nrow(a_valid_data)) * 100` percent of the study is women participants, and `r (nrow(filter(a_valid_data, sex == "Female", apoe4 == TRUE)) / nrow(filter(a_valid_data, sex == "Female"))) * 100` percent of women in the study are APOE4 carriers.


Next, we can import the biomarker dataset.

```{r}
b_data = 
  read_csv("data/data_mci/mci_amyloid.csv",
           skip = 1,
           na = c("NA", "Na")) |> 
  janitor::clean_names() |> 
  rename(time_0 = baseline) |> 
  pivot_longer(
    time_0:time_8,
    names_to = "years_elapsed",
    names_prefix = "time_",
    values_to = "amyloid_ratio"
  ) |> 
  mutate(
    across(years_elapsed, as.double),
    across(amyloid_ratio, as.double)
  )

```

First, we import the data, skip the first row since its descriptive and not relevant, and mark "NA" and "Na" as the missing value. Then we can clean names to make them consistent and easy to use. Next, we can look at our data and see that the column names `r names(b_data)[1:5]` are observed values, where the baseline and biomarker measurements are described as "Time (in years) elapsed since the study baseline to the visit where biomarker Amyloid _ 42/40 ratio was measured". Subsequently, we use the `pivot_longer` function to transform these into two columns, years elapsed and amyloid ratio. Before that, we also want to change the baseline variable to time_0, so it aligns with the other column names that we want to transform. Lastly, we change the all values from characters to doubles so that they are mathematically interpretable.

The cleaned dataset has `r nrow(b_data)` participants (rows) with `r names(b_data)` as the data points we have for each participants (cols). For the years_elapsed column, we have values `r unique(pull(b_data, years_elapsed))` for each participant, tracking the amyloid ratio through the study. For example, we have baseline/years elapsed equal to 0 with a mean value of `r mean(!is.na(pull(filter(b_data, years_elapsed == 0), amyloid_ratio)))`.

Now, we can check the number of participants who appear only in the baseline or only in the amyloid dataset. To accomplish this, we can use the `left_join` function to join the baseline dataset to the amyloid dataset by the id/study id variables. Once they are joined, for participants who only appear in the baseline, we can just check if they have NA values for the amyloid ratio, which would mean that they are not in the amyloid dataset. We can similarly check who is only in the Amyloid dataset by switching the order of datasets in `left_join` and changing the criteria to whether education exists (or any other variable in baseline dataset that does not exist in amyloid dataset).

Note, since we originally changed the amyloid dataset to have time set as one column, there are ids with multiple instances for each elapsed year, so here, we use the pre-pivoted dataset since there is only one row per id. Additionally, for the participants who only were in the amyloid dataset, we generically chose education as the field to indicate whether someone was in the baseline dataset, so we do an additional check at the end to make sure there were no NA values in education for those in the baseline dataset, which results in zero.

```{r}
b_data_original = 
  read_csv("data/data_mci/mci_amyloid.csv",
           skip = 1,
           na = c("NA", "Na")) |> 
  janitor::clean_names() |> 
  rename(time_0 = baseline)

unique_a =
  left_join(a_valid_data, b_data_original, by = c("id" = "study_id")) |> 
  filter(if_all(starts_with("time_"), is.na)) |> 
  distinct()

unique_b =
  left_join(b_data_original,a_valid_data , by = c("study_id" = "id")) |> 
  filter(is.na(education)) |> 
  distinct()

cat("The number of participants only in baseline dataset is", nrow(unique_a))
cat("The number of participants only in amyloid dataset is", nrow(unique_b))
cat("The number of participants in baseline dataset with missing education is", nrow(filter(a_valid_data, is.na(education))))
```

As mentioned in the output code, we can see that there are 8 participants only in the baseline dataset and 15 participants only in the amyloid dataset. Of those only included in the baseline, only `r nrow(filter(unique_a, !is.na(age_at_onset)))` developed MCI and of those only included in the amyloid dataset, only `r nrow(filter(unique_b, is.na(time_0)))` did not record a baseline amyloid ratio.

Now looking at the participants who were in both baseline and amyloid datasets, we can build the entire combined dataset and then remove those who's ids align with the ones we just found to not be in either dataset. Since the `left_join` function will already remove those ids not in baseline, all we need to do is remove those that are only in baseline using `filter`.

```{r}
combined_data = 
  left_join(a_valid_data, b_data, by = c("id" = "study_id")) |> 
  filter(!id %in% pull(unique_a, id))

```

The resulting dataset has `r nrow(combined_data)` data points (rows), where each datapoint is a participant at a specific elapsed year, and it has `r ncol(combined_data)` variables per observation (cols). In this dataset, we have only included participants of both the baseline and amyloid datasets. At the baseline year of those who were measured, they have an average age of `r round(mean(pull(filter(combined_data, years_elapsed == 0), current_age), na.rm = TRUE), 2)` years old and average amyloid ratio of `r round(mean(pull(filter(combined_data, years_elapsed == 0), amyloid_ratio), na.rm = TRUE), 4)`.

```{r}
write_csv(combined_data, "data/combined_data.csv")
```












