---
title: "p8105_hw2_fs2968"
output: github_document
date: "2026-09-30"
---
Importing necessary libraries
```{r}
library(tidyverse)
```


## Problem 1, NYC Transit Data

```{r}
t_data = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude, starts_with("route"), entry, vending, entrance_type, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```


In the above code, we read in the NYC transit subway data-set that includes information about the following points: line, station name, station latitude and longitude, the route it lies on, whether there is an entry and whether there is a fee, and the entrance type. To clean the data, we first use the `clean_names` function from the janitor package to ensure all of the column names are cleaned so that there are no dots, uppercase, and overall more consistent. The next step was selecting all of the columns that we want to keep using the `select` function applied on the imported data. Note, since we wanted all route information, we use the `starts_with` function to include all columns with names that start with route to simplify the code. Lastly, we use the `mutate` function to change the data-set, modifying the entry column to be a logical vector with `TRUE` and `FALSE` instead of characters `"YES"` and `"NO"`. In the import, you could consider declaring `NA= c("NA", "", ".")` a cleaning step, but it is more of an identification for the read_csv function of what the missing values will be. 

The post-clean data set has dim (`r dim(t_data)`), which represents `r dim(t_data)[1]` transit stations (rows) with `r dim(t_data)[2]`  descriptor fields per row (columns). After all of the cleaning, the data is mostly tidy. The names look good and data typing is in line with normal standards, but one thing we are missing is that some of the column names are observation values, specifically route1,...,route11. Instead of separating these into 11 different groups, we should instead have this as one column for routes.

To find the number of distinct stations, which vary by name and line, we can use the `distinct` and `nrow` functions:

```{r}
nrow(distinct(t_data, station_name, line))
```

To find the number of stations that are ADA compliant, we look at all stations that are ADA compliant, and then add the `distinct` function to ensure we aren't double counting any stations, and then use `nrow` to find the total.

```{r}
filter(t_data, ada == TRUE) |> 
  distinct(station_name, line) |> 
  nrow()
```

To find the number of station entrances/exits without vending that allow entrance, we first `filter` t_data for those that allow entrance and those without vending. Once we have that filtered data-set, we similarly use `distinct` to make sure we aren't repeating stations and use `nrow` to find the total. Now we have all of the pieces and we simply divide that value by the total number of distinct stations without vending to find the proportion:

```{r}
numerator = filter(t_data, entry == TRUE, vending == "NO") |> 
            distinct(station_name, line) |> 
            nrow()

denominator = filter(t_data, vending == "NO") |> 
              distinct(station_name, line) |> 
              nrow()

numerator/denominator
```


To reformat the data, we will use the `pivot_longer` function to change the route columns to describe the route number as one column and the route name as another. Before we do that, we first need to change the route names to all be char because route 1,2,3,4,... are default to doubles:

```{r}
reformatted_t_data = 
  t_data |> 
  mutate(
    across(starts_with("route"), as.character)
  ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )

```
Now we can answer the find the number of distinct stations that serve the A train by using the `filter` function to only look at stations that serve the A, and then use the `distinct` and `nrow` functions to find the total distinct stations:

```{r}
filter(reformatted_t_data, route_name == "A") |> 
  distinct(station_name, line) |> 
  nrow()
```

To further find the number that are ADA compliant, we simply add that to the previous `filter` function:

```{r}
filter(reformatted_t_data, route_name == "A", ada == TRUE) |> 
  distinct(station_name, line) |> 
  nrow()
```


## Problem 2, Trash Wheel Dataset

We first read in the Trash Wheel Data-set and remove non-data entries and not including dumpster specific data. We also can adjust the names to be more reasonable and round the number of sports balls to the nearest integer and conver to an interger type using `as.integer`.

```{r}
trash_data = 
  readxl::read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx", 
    sheet="Mr. Trash Wheel", 
    range="A2:N709",
    na = c("#", "")) |> 
  janitor::clean_names() |> 
  mutate(
    across(sports_balls, round),
    across(sports_balls, as.integer),
    owner = "Mr. Trash",
    across(year, as.double)
  )

```
Now we can grab the other two datasets and clean them using a similar process:

```{r}
prof_data = 
  readxl::read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx", 
    sheet="Professor Trash Wheel", 
    range="A2:M134",
    na = c("#", ""))|> 
  janitor::clean_names() |> 
  mutate(
    owner = "Professor Trash",
    )

gwyn_data = 
  readxl::read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx", 
    sheet="Gwynns Falls Trash Wheel", 
    range="A2:L351",
    na = c("#", ""))|> 
  janitor::clean_names() |> 
  mutate(owner = "Gwynns Trash")

```


Now we can use the `bind_rows` function to join these data-sets together. Before that, we can create a column in all three called "owner", which designates who the data comes from, editted in the above code blocks. 

```{r}
total_trash = bind_rows(trash_data, prof_data, gwyn_data)
```

The resulting data-set `total_trash` is the combination of all three excel sheets. This combined data-set includes `r nrow(total_trash)` observations (rows) and `r ncol(total_trash)` variables (columns). There are both identification variables and descriptive variables. The identification variables include `r names(total_trash)[0:4]`, which help us identify the dumpsters and when the were observed. The descriptive variables include `r names(total_trash)[5:ncol(total_trash)]`, which describe the weight, volume, and breakdown of what is found in the dumpster. The weight is measured in tons with mean value of `r mean(pull(total_trash, weight_tons))`, and the volume is measured in cubic yards with an mean value of `r mean(pull(total_trash, volume_cubic_yards)) `. The other descriptive variables that detail what actual items are in the dumpters use integer interpretations, for example, the first 3 dumpsters had `r pull(total_trash, plastic_bags)[0:3]` plastic bags.

For the available data, the total weight of trash collected by Professor Trash Wheel was `r sum(pull(total_trash, weight_tons))` tons. We can additionally find the total number of cigarette butts collected by Gwynnda in June of 2022 as `r sum(pull(filter(total_trash, owner == "Gwynns Trash", month == "June", year == 2022), cigarette_butts))`.


## Problem 3, Alzheimer's Disease Dataset











